On Optimality of Finite Element Integration

### 2016-01-14 TOMS_Paper/assembly.tex
Several fixes, but very important

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality and quasi-optimality are defined in terms of floating point operations given a memory bound. We provide an approach to explore the space of legal transformations and discuss under what conditions optimality or quasi-optimality hold. A theoretical analysis and extensive experimentation, which shows systematic performance improvements over several state-of-the-art code generation systems, validate the approach.
\end{abstract}

@@

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (``quasi-optimality'') in all other cases. Summarizing, our contributions are as follows:

\begin{itemize}
\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We explain under what conditions our algorithm leads to optimality. In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, in general, always achieved (i.e., the operation count is at least optimal in innermost loops).
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. Expressions (right hand side of a statement) result from the translation of a form in high level matrix notation into code. In particular, $m$ is the number of ``terms'' (a monomial can be implemented by one or more terms), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex. We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.). This bring us to formalize the class of loop nests for which we seek optimality.

@@

\begin{Def}[Finite element integration loop nest]
\label{def:fem-loopnest}
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which each statement has expressions in normal form.
\end{Def}

@@

\begin{Def}[Sharing]
A statement within a loop nest $\Lambda$ presents sharing if at least one of the following conditions hold:
\begin{enumerate}
\item there are at least two symbolically identical sub-expressions (spatial sharing)
\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
\end{enumerate}
\end{Def}

To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest. In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$. Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}). Finally, generalized code motion \cite{Luporini} leads to optimality (Figure~\ref{code:multi_loopnest_c}).

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold. Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

Section~\ref{sec:se-rln} discusses structural and algebraic properties characterizing our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. Examples are provided in Section~\ref{sec:se-examples}. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For example, suppose that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion -- not needed in the example as we have one loop.
\item Searching for spatial sharing, expanding and factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on multiple factors: the loop size, the expansion cost, and code motion. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any tensor operation that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

Of notable importance, from the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations. Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest. It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration. Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality and quasi-optimality are defined in terms of floating point operations given a memory bound. We provide an approach for exploring the space of legal transformations, which systematically reaches quasi-optimality. The conditions under which optimality is achieved are also discussed. A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
\end{abstract}

@@

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (we will define ``quasi-optimality'', which relates operation count with inner loops) in all other cases. Summarizing, our contributions are as follows:

\begin{itemize}
\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We demonstrate that our algorithm systematically leads to quasi-optimality. We also explain under what conditions of the input problem optimality is achieved.
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. Expressions (right hand side of a statement) result from the translation of a form in high level matrix notation into code. In particular, $m$ is the number of ``terms'' (a monomial can be implemented by one or more terms), $\alpha_{eij}$ ($\beta_{eik}$) represents the product of a coefficient function (e.g., the inverse Jacobian matrix for the change of coordinates) with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We do not pose any restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex. We say that such an expression is in \textit{normal form}, because the algebraic structure of a variational form is intact (e.g., products have not been expanded yet, distinct monomials can still be identified, etc.). This bring us to formalize the class of loop nests for which we seek optimality.

@@

\begin{Def}[Finite element integration loop nest]
\label{def:fem-loopnest}
A finite element integration loop nest is a loop nest in which appear, in order, an imperfect free order loop, an imperfect (perfect only in special cases), linear or non-linear reduction loop, and a multilinear loop nest whose body is an independent basic block in which all expressions are in normal form.
\end{Def}

@@

\begin{Def}[Sharing]
A statement within a loop nest $\Lambda$ presents sharing if at least one of the following conditions hold:
\begin{enumerate}
\item there are at least two symbolically identical sub-expressions (spatial sharing)
\item there is at least one non-trivial sub-expression (an addition or a product) that is redundantly executed, as independent of $\lbrace L_{i_0}, L_{i_1}, ...L_{i_n} \rbrace \subset \Lambda$ (temporal sharing)
\end{enumerate}
\end{Def}

To illustrate the definition, we show in Figure~\ref{code:multi_loopnest} how sharing evolves as factorization and code motion are applied to a trivial multilinear loop nest. In the original loop nest (Figure~\ref{code:multi_loopnest_a}), spatial sharing is induced by $b_j$. Factorization eliminates spatial sharing and promotes temporal sharing (Figure~\ref{code:multi_loopnest_b}). Finally, generalized code motion \cite{Luporini}, which seeks redundant sub-expressions analyzing all levels of the loop nest\footnote{Traditional loop-invariant code motion, which is commonly applied by general-purpose compilers, only considers innermost loops.}, leads to optimality (Figure~\ref{code:multi_loopnest_c}).

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and the expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This requires coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold. Setting quasi-optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

%Section~\ref{sec:se-rln} discusses structural and algebraic properties characterizing our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. Examples are provided in Section~\ref{sec:se-examples}. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form.

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, consider the tensor operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$. To schedule sub-expressions of this kind, we have two options:
\begin{enumerate}[label=(\roman*)]
\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$. In general, however, the optimal choice depends on multiple factors: the loop size, the increase in operation count due to expansion, and the gain in code motion. We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii). Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured; this observation is fundamental for understanding the sharing elimination algorithm, described in the next section\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}.

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes how we perform sharing elimination assuming as input a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest. Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.
